Generating Random Logic Programs Using Constraint Programming

### 2020-01-09 paper/paper.tex
Some writing on counting

\section{Heads of Clauses}
% TODO: section for clauses. define heads & bodies, then separate subsections
% for head and body constraints

@@

\section{Bodies of Clauses}

@@

\subsection{Nodes}

@@

\subsection{Constraints}

@@

\section{Eliminating Variable Symmetries} \label{sec:variable_symmetry}

Given any clause, we can permute the variables in it without changing the
meaning of the clause or the entire program. Thus, we want to fix an order on
variables to eliminate unnecessary symmetries. Informally, we can say that
variable $X$ goes before variable $Y$ if its first occurrence in either the head
or the body of the clause is before the first occurrence of $Y$.

\begin{definition}
  Head:

  size: $|\constants{}| + |\variables{}|$

  domain: $\{ 0, \dots, \maxArity{} - 1\}$

  condition: for $i = 0, \dots, \maxArity{} - 1$ and $t \in \constants{} \cup \variables{}$,
  \[
    \variable{arguments}[i] = t \quad \iff \quad i \in \variable{occurrences}[t].
  \]

  Body:

  size: $|\constants{}| + |\variables{}| + 1$

  domain: $\maxNumNodes{} \times \maxArity{}$

  condition: for $i = 0, \dots, \maxNumNodes{} - 1$, $j = 0, \dots, \maxArity{}
  - 1$, $t \in \constants{} \cup \variables{} \cup \{ \Box \}$,
  \[
    \variable{treeValues}[i].\variable{arguments}[j] = t \quad \iff \quad i
    \times \maxNumNodes{} + j \in \variable{occurrences}[t]
  \]

  We define an array \variable{occurrences} of  sets such that, for all $t \in
  \constants{} \cup \variables{}$, $\variable{occurrences}[t]$ is a subset of
  that stores the indices of \variable{arguments} at which $t$ is located. It is
  defined by the \variable{setsIntsChanneling} constraint equivalent to the
  following condition:
\end{definition}

\begin{definition}
  We define a $|\variables{}| \times 2$ matrix $\mathbf{M}$ with domain $[0,
  \maxNumNodes{} \times \maxArity{}]$ such that for $v = 0, \dots, |\variables{}| - 1$,
  \begin{align*}
    M_{v,0} &= |\variable{occurrences}[v]|.\\
    M_{v,1} &= \begin{cases}
      \min \variable{occurrences}[v] & \text{if } \variable{occurrences}[v] \ne \emptyset\\
      \maxNumNodes{} \times \maxArity{} & \text{otherwise.}
    \end{cases}
  \end{align*}
\end{definition}

\begin{constraint}
  We can then impose the \variable{lexChainLessEq} constraint on $\mathbf{M}$
  which means that for $v = 1, \dots, |\variables{}| - 1$,
  \[
    |\variable{occurrences}[v - 1]| \le |\variable{occurrences}[v]|,
  \]
  and if $|\variable{occurrences}[v - 1]| = |\variable{occurrences}[v]| > 0$,
  then
  \[
    \min \variable{occurrences}[v - 1] \le \min \variable{occurrences}[v].
  \]
\end{constraint}

\begin{example}
  Let $\maxArity{} = 4$, $\mathsf{P} \in \predicates{}$, $\constants{} = \{ a
  \}$, $\variables{} = \{X, Y, Z \}$. Then $\mathsf{P}(Z, Y, Z)$ would be
  represented as:
  \begin{align*}
    \variable{predicate} &= \mathsf{P},\\
    \variable{arguments} &= [Z, Y, Z, 0],\\
    \variable{arity} &= 3,\\
    \variable{occurrences} &= [\emptyset, \emptyset, \{ 1 \}, \{ 0, 2 \}],\\
    \mathbf{M} &= \begin{bmatrix}
      0 & 0\\
      1 & 1\\
      2 & 0
    \end{bmatrix}.
  \end{align*}
\end{example}

@@

\section{Counting Programs}

This is without any kind of cycle detection and without probabilities.

Number of ways to fill $n$ positions with terms:
\[
  P(n) = |\constants{}|^n + \sum_{\substack{1 \le k \le |\variables{}|, \\ 0 =
      s_0 < s_1 < \dots < s_k < s_{k+1} = n+1}} \prod_{i=0}^k (|\constants{}| +
  i)^{s_{i+1} - s_i - 1}
\]

Let $p_a$ be the number of predicates in $\predicates{}$ with arity $a \in
\arities{}$. Number of clauses with total arity $a$:
\begin{align*}
  C(0) &= 1,\\
  C(a) &= \sum_{n=1}^{\maxNumNodes{}} T(n, a),
\end{align*}
where $T(n, a)$ is defined recursively as:
\[
  T(1, a) = p_a
\]
and
\[
  T(n, a) = T(n-1, a) + 2\sum_{\substack{c_1 + \dots + c_k = n - 1,\\
      2 \le k \le \frac{a}{\min \arities{}},\\
      c_i \ge 1 \text{ for all } i}} \sum_{\substack{d_1 + \dots + d_k = a,\\
    d_i \ge \min \arities{} \text{ for all } i}} \prod_{i=1}^k T(c_i, d_i).
\]

Let us order the elements of $\predicates{}$, and let $a_i$ be the arity of the
$i$-th predicate. The number of programs is then:
\[
  \sum_{\substack{ \sum_{i=1}^{|\predicates{}|} h_i = n,\\
      |\predicates{}| \le n \le \maxNumClauses,\\
      h_i \ge 1 \text{ for all } i}} \prod_{i=1}^{|\predicates{}|}
  \multiset{\sum_{a=0}^{\maxArity{} \times \maxNumNodes{}} C(a) P(a+a_i)}{h_i},
\]
where
\[
  \multiset{n}{k} = \binom{n+k-1}{k}.
\]

>>>>

\section{Considering Each Clause Separately}
% TODO: define heads & bodies

\subsection{Heads of Clauses}
% TODO: section for clauses. define heads & bodies, then separate subsections
% for head and body constraints

@@

\subsection{Bodies of Clauses}

@@

\subsubsection{Nodes}

@@

\subsubsection{Constraints}

@@

\subsection{Eliminating Variable Symmetries} \label{sec:variable_symmetry}

Given any clause, we can permute the variables in it without changing the
meaning of the clause or the entire program. Thus, we want to fix the order of
variables to eliminate unnecessary symmetries. Informally, we can say that
variable $X$ goes before variable $Y$ if its first occurrence in either the head
or the body of the clause is before the first occurrence of $Y$.

\begin{definition}
  % TODO: what odes this mean? example?
  Let $N = \maxArity{} \times (\maxNumNodes{} + 1)$. Let
  $\variable{termsPerClause}[N] \in \constants{} \cup \variables{} \cup \{ \Box
  \}$ be a matrix, each row of which is a flattened list of all positions for
  terms in a particular clause.

  Then we can use the $\variable{setsIntsChanneling}$ constraint
  to define $\variable{occurrences}[|\constants{}| + |\variables{}| + 1]$ as a
  matrix of subsets of $\{ 0, \dots, N-1 \}$ such that for all $i = 0, \dots, N
  - 1$, and $t \in \constants{} \cup \variables{} \cup \{ \Box \}$,
  \[
    i \in \variable{occurrences}[t] \quad \iff \quad
    \variable{termsPerClause}[i] = t
  \]
\end{definition}

\begin{definition}
  We define $\arrayd{M}{|\variables{}|}{\{ 0, \dots, N \}}$ such that for $v \in
  \variables{}$,
  \[
    \variable{M}[v] = \begin{cases}
      \min \variable{occurrences}[v] & \text{if }
      \variable{occurrences}[v] \ne \emptyset\\
      N & \text{otherwise.}
    \end{cases}
  \]
\end{definition}

\begin{constraint}
  $\variable{M}$ is sorted.
\end{constraint}

\begin{example} % TODO: update the example to cover an entire clause
  Let $\maxArity{} = 4$, $\mathsf{P} \in \predicates{}$, $\constants{} = \{ a
  \}$, $\variables{} = \{X, Y, Z \}$. Then $\mathsf{P}(Z, Y, Z)$ would be
  represented as:
  \begin{align*}
    \variable{predicate} &= \mathsf{P},\\
    \variable{arguments} &= [Z, Y, Z, 0],\\
    \variable{arity} &= 3,\\
    \variable{occurrences} &= [\emptyset, \emptyset, \{ 1 \}, \{ 0, 2 \}],\\
    \mathbf{M} &= \begin{bmatrix}
      0 & 0\\
      1 & 1\\
      2 & 0
    \end{bmatrix}.
  \end{align*}
\end{example}

@@

\section{Counting Programs}
% TODO: define combined arity.
% TODO: define 'gap', 'argument', 'position' or something like that.

In order to demonstrate the correctness of the model and explain it in more
detail, in this section we are going to derive combinatorial expressions for
counting the number of programs with up to $\maxNumClauses{}$ clauses and up to
$\maxNumNodes{}$ nodes per clause, and arbitrary $\predicates{}$,
$\arities{}$, $\variables{}$, and $\constants{}$. To simplify the task, we only
consider clauses without probabilities and disable (negative) cycle elimination.
It was experimentally confirmed that the model agrees with the combinatorial
formula from this section in 985 different scenarios.

We will first consider clauses with `gaps', i.e., without taking variables and
constants into account. Let $T(n, a)$ denote the number of possible clause
bodies with $n$ nodes and combined arity $a$. Then $T(1, a)$ is the number of
predicates in $\predicates{}$ with arity $a$, and the following recursive
definition can be applied for $n > 1$:
\[
  T(n, a) = T(n-1, a) + 2\sum_{\substack{c_1 + \dots + c_k = n - 1,\\
      2 \le k \le \frac{a}{\min \arities{}},\\
      c_i \ge 1 \text{ for all } i}} \sum_{\substack{d_1 + \dots + d_k = a,\\
    d_i \ge \min \arities{} \text{ for all } i}} \prod_{i=1}^k T(c_i, d_i).
\]
The first term here represents negation, i.e., negating an expression consumes
one node but otherwise leaves the task unchanged. If the first operation is not
negation, then it must be either conjunction or disjunction (hence the
coefficient `2'). In the first sum, $k$ represents the number of children of the
root node, and each $c_i$ is the number of nodes dedicated to child $i$. Thus,
the first sum iterates over all possible ways to partition the remaining $n-1$
nodes. Similarly, the second sum considers every possible way to partition the
total arity $a$ across the $k$ children nodes.

We can then count the number of possible clause bodies with total arity $a$ (and
any number of nodes) as
\[
  C(a) = \begin{cases}
    1 & \text{if } a = 0\\
    \sum_{n=1}^{\maxNumNodes{}} T(n, a) & \text{otherwise.}
  \end{cases}
\]
Here, the empty clause is considered separately.

The number of ways to fill $n$ positions with terms can be expressed as
\[
  P(n) = |\constants{}|^n + \sum_{\substack{1 \le k \le |\variables{}|, \\ 0 =
      s_0 < s_1 < \dots < s_k < s_{k+1} = n+1}} \prod_{i=0}^k (|\constants{}| +
  i)^{s_{i+1} - s_i - 1}.
\]
The first term is simply the number of ways to fill all $n$ positions with
constants. The parameter $k$ is the number of variables used in the clause, and
$s_1, \dots, s_k$ mark the first occurrence of each variable. For each gap
between any two introductions (or before the first introduction, or after the last
introduction), we have $s_{i+1}-s_i-1$ spaces to be filled with any of the
$\constants{}$ constants or any of the $i$ already-introduced variables.

Let us order the elements of $\predicates{}$, and let $a_i$ be the arity of the
$i$-th predicate. The number of programs is then:
\[
  \sum_{\substack{ \sum_{i=1}^{|\predicates{}|} h_i = n,\\
      |\predicates{}| \le n \le \maxNumClauses,\\
      h_i \ge 1 \text{ for all } i}} \prod_{i=1}^{|\predicates{}|}
  \multiset{\sum_{a=0}^{\maxArity{} \times \maxNumNodes{}} C(a) P(a+a_i)}{h_i},
\]
where
\[
  \multiset{n}{k} = \binom{n+k-1}{k}
\]
counts the number of ways to select $k$ out of $n$ items with repetition (and
without ordering). Here, we sum over all possible ways to distribute
$|\predicates{}| \le n \le \maxNumClauses{}$ clauses among $|\predicates{}|$
predicates so that each predicate gets at least one clause. For each predicate,
we can then count the number of ways to select its clauses out of all possible
clauses. The number of possible clauses can be computed by considering each
possible arity $a$, and multiplying the number of `unfinished' clauses $C(a)$ by
the number of ways to fill the $a+a_i$ gaps in the body and the head of the
clause with terms.
